Formal Grammars as Types in Non-Commutative Linear-Non-Linear Type Theory

### 2024-07-11 POPL/paper.tex
Some ideas on themes of paper + spelling

  To demonstrate the usefulness of the calculus for formal grammar
  theory, we show that existing grammar formalisms --- e.g.\ regular
  expressions, context free expressions, and their associated automata \max{which ones do
    we actually show?} \steven{regular expressions, $\mu$-regular expressions
    (or context free expressions), and different automata classes (finite,
    pushdown, turing)} --- can be expressed as linear types in
  \theoryabbv. We give simple characterizations of the regular and
  context-free grammars as expressible by restricting which
  connectives of \theoryabbv are allowed, a form of type theoretic
  characterization of these grammar classes. Further, we demonstrate
  some central proofs of formal grammar theory can be intrinsically
  verified: \emph{strong} equivalence between regular expressions and
  NFAs is expressed by a syntactic \emph{isomorphism} definable in
  \theoryabbv.

@@

Parsing data from a serialized format is one of the most common tasks
in all of computing. Accordingly, this is a quite well-studied problem
spanning theoretical computer science and linguistics with such
milestones as Chomsky's hierarchy of grammars
\cite{chomThreeModels1956}, practical algorithms for parsing of
regular and context-free grammars, and variants
\cite{KNUTH1965607,Earley1970}, as well as implementations of
practical tools for the generation of efficient parsers
\cite{Johnsonyacc}.
%
The final frontier for parsing research is the development of
\emph{formally verified} parser implementations, which due to the
pervasive nature of parsing would be a useful component of many
formally verified software systems: compilers and servers most
especially.
%
\steven{Some ideas on what the unifying message throughout the paper should be:

}

In this work we aim to refine formal language theory into a
mathematical theory of formal grammars that can be used to develop
\emph{correct-by-construction} parsers. Our approach can be summarized
in the Curry-Howard-style slogan ``grammars as types'', and the
addendum ``parsers as terms'' where since the types of the terms are
grammars, this means not just parsers but intrinsically verified
parsers as terms. That is, we develop a type theory with a novel
denotational semantics in a category with abstract formal grammars as
objects and \emph{parse transformers} as morphisms. We say these are
\emph{abstract} formal grammars because they are not tied to any
particular syntactic formalism such as Chomskyian grammars, regular
expressions, Lambek calculus etc. The category we choose is a rich
category: it is a topos as well as being monoidal closed, and these
universal properties serve as the inspiration for our type theory.

@@

\paragraph{Contributions} This paper begins in \cref{sec:synindnotion} by
giving a universal semantic characterization of formal grammar.
From there, in \cref{sec:tt} we provide the syntax of \theoryname, a type thoery
that axiomatizes the underlying structures needed to conduct formal grammar
theory. Then in \cref{sec:categorify}, we characterize the abstract categorical
models of our syntax. Afterwards, we demonstrate several models of \theoryabbv,
including one to prove a metatheoretic canonicity theorem. Finally, in
\cref{sec:automata} we give a presentation of automata as grammars. With this
view in mind, we may encode parsing algorithms and execute them entirely
internal to our formalism. Our contributions are then:

@@

Unfortunately, programming in a linear term language is often
cumbersome and unintuitive for users who are unacustomed to the usage
restrictions.\max{You are arguing against our own type theory here...} \steven{I
  want to argue against a \emph{purely linear} calculus to motivate the
  convience of LNL, but perhaps this isn't appropriate here, and I don't want to
  make our own type theory sound bad.Should probably just cut the beginning of
  this paragraph and jump into LNL, but then there's not an immediate reason for
why one might want/need it over a pure linear system} Code in such a language can
become unnecessarily verbose when ensuring every function written is indeed linear. It would be more ergonomic to allow users using Girard encodings to translate
high-level reasoning into low-level linear terms.\pedro{What encodings are these?
We should add a citation here}
To alleviate this
concern, in 1995 Benton et al.\ proposed an alternative categorical
semantics of linear logic based on adjoint interactions between linear
and non-linear logics \cite{bentonMixedLinearNonlinear1995} ---
appropriately referred to as a \emph{linear-non-linear} (LNL)
system. This work is simply typed, so the boundary between linear and
non-linear subtheories is entirely characterized via a monoidal
adjunction $F \dashv G$ between linear terms and non-linear terms.

@@

\pedro{The paragraphs above are a bit circular, what about the following alternative?}
\steven{I am in favor of cutting what's above, but we haven't introduced the
  sytnax in the main body of the paper yet, even if the figures appeared
  earlier. So what you write needs to take that into account}

@@

\paragraph{Regular Expressions} We may realize regular expression as a subtheory
of our language by restricting the type constructors to the
linear unit, characters, $\otimes$, $\oplus$, and Kleene star. Classically, the
languages recognized by these are exactly the regular languages --- the lowest
on the Chomksy hierarchy.

@@

Though Kleene categories are expressive enough to reason about
concepts that are outside of reach of Kleene algebras, their
simply-typed nature makes them not so expressive from a type theoretic
point of view. This is limiting because type theories are sucessful
syntactic frameworks for manipulating complicated categorical
structures while avoiding some issues common in category theory, such
as coherence issues.

@@

\steven{Max suggests augmenting the definition of a chomsky category to
  something like two categories $L$ and $U$, and $U$ is cartesian closed I don't
  fully recall the rest.

@@

\pedro{This is likely true, but if we explicitly say so, this warants a proof. I think that
if we don't say anything about the syntactic category, reviewers won't mind.}

\steven{I agree that we don't want to say anything that opens unnecessary
  questions for proofs we haven't written. However, it seems hard to make the
  case that we have the right categorical model of the syntax if this statement
  isn't true. By restating our definition of Chomksy category, this should be
  obvious or a quick proof}

@@

Further, semantic actions can be arranged into a structured category.
Define $\SemAct$ as the comma
category $\Grammar / \Delta$, where $\Delta : \Set \to \Grammar$ defines a
discrete presheaf. That is, for a set $X$, $\Delta (X)(w) = X$ for all
$w \in \String$. As $\SemAct$ is defined as a comma category, it has a forgetul
functor into $\Grammar$. That is, $\SemAct$ serves as a notion of formal
grammar. Moreover, $\SemAct$ is a model of \theoryabbv.

@@

The canonicity theorem from the previous section gives a tight connection
between terms in our type theory and the theory of formal languages. In this section
we further explore these connections by using the fact that, clasically
formal language theory is closely related to the study of automata.
In the Chomsky hierarchy, each language class is associated to a class of
automata that serve as recognizers. Internal to \theoryabbv,
we can characterize these language classes syntactically; moreover, we
demonstrate the equivalence of these language classes to its associated automata class as a
proof term within our logic.

@@

In 1992, while investigating the addition of arbitrary recursion to Kleene
algebra \cite{leiss},
Lei{\ss} used a least fixed-point operator instead of the Kleene star. Although
he did not explicitly make use of a type system, or have access to the view of Frisch and
Cardelli, Leiss had added the full power of inductive types to a type system of regular expressions.

@@

\steven{Mention Neel's typed algebraic approch paper here. Say it is similar at
  first glance, as its parsing with types, but that he actually is representing
  something quite different. Or is it so different that its not worth mentioning?}

@@

\steven{Can we sketch here what an apt categorification of KATs might be? As a
  poor first pass: perhaps a Kleene category with a full involutive
  subcategory that corresponds to the Boolean aglebra?.}
\steven{What can we say about internalizing Kleene algebra proofs or extensions?}
\steven{any KA citations needed above?}

@@

\cite{fu2023twolevellineardependenttype} give a compiler for a two-level
linear dependent type theory. This could serve as a template for a
compiler implementation of \theoryname that is not just inference rules
embedded in agda.

\paragraph{Implementations in Parsing}
Talk about how we have a DFA parser in agda and want to to get out a regex
parser. Then context free, etc

@@

\paragraph{Other Future Work}
Backtracking parsing via delimited continuations?

>>>>

  To demonstrate the usefulness of the calculus for formal grammar
  theory, we show that existing grammar formalisms --- e.g.\ regular
  expressions, context free expressions, and their associated automata \max{which ones do
    we actually show?} \steven{regular expressions, $\mu$-regular expressions
    (or context free expressions), and different automata classes (finite,
    pushdown, Turing)} --- can be expressed as linear types in
  \theoryabbv. We give simple characterizations of the regular and
  context-free grammars as expressible by restricting which
  connectives of \theoryabbv are allowed, a form of type theoretic
  characterization of these grammar classes. Further, we demonstrate
  some central proofs of formal grammar theory can be intrinsically
  verified: \emph{strong} equivalence between regular expressions and
  NFAs is expressed by a syntactic \emph{isomorphism} definable in
  \theoryabbv.

@@

Parsing data from a serialized format is one of the most common tasks
in all of computing. Accordingly, this is a quite well-studied problem
spanning theoretical computer science and linguistics with such
milestones as Chomsky's hierarchy of grammars
\cite{chomThreeModels1956}, practical algorithms for parsing of
regular and context-free grammars, and variants
\cite{KNUTH1965607,Earley1970}, as well as implementations of
practical tools for the generation of efficient parsers
\cite{Johnsonyacc}.
%
The final frontier for parsing research is the development of
\emph{formally verified} parser implementations, which due to the
pervasive nature of parsing would be a useful component of many
formally verified software systems: compilers and servers most
especially.
%
\steven{Some ideas on what the unifying message throughout the paper should be:
we offer an abstraction over the formalisms used ad-hoc when doing formal
language/grammar theory. Whether a user of these constructs recognizes it or not, there is some
universality between all of their systems. We don't necessarily do much, per se, other
than abstract over the commonalities and provide a common language.

However, just because there isn't a ton of clever work in designing this syntax, as
a lot of it does flow very naturally from the literature, that doesn't diminish
the value of the syntax. In fact, it emboldens the syntax because we want to
appeal to the idea that ``there couldn't have been any other choice for
universal grammar''.

That is all to say, our appeal is generalization. I think does appear
throughout the paper, but is maybe too aimless of a message to be satisfying.
Perhaps, we should also lean on applications more. Idk I'm gonna sleep now, and
we'll talk in the morning
}

In this work we aim to refine formal language theory into a
mathematical theory of formal grammars that can be used to develop
\emph{correct-by-construction} parsers. Our approach can be summarized
in the Curry-Howard-style slogan ``grammars as types'', and the
addendum ``parsers as terms'' where since the types of the terms are
grammars, this means not just parsers but intrinsically verified
parsers as terms. That is, we develop a type theory with a novel
denotational semantics in a category with abstract formal grammars as
objects and \emph{parse transformers} as morphisms. We say these are
\emph{abstract} formal grammars because they are not tied to any
particular syntactic formalism such as Chomskyan grammars, regular
expressions, Lambek calculus etc. The category we choose is a rich
category: it is a topos as well as being monoidal closed, and these
universal properties serve as the inspiration for our type theory.

@@

\paragraph{Contributions} This paper begins in \cref{sec:synindnotion} by
giving a universal semantic characterization of formal grammar.
From there, in \cref{sec:tt} we provide the syntax of \theoryname, a type theory
that axiomatizes the underlying structures needed to conduct formal grammar
theory. Then in \cref{sec:categorify}, we characterize the abstract categorical
models of our syntax. Afterwards, we demonstrate several models of \theoryabbv,
including one to prove a metatheoretic canonicity theorem. Finally, in
\cref{sec:automata} we give a presentation of automata as grammars. With this
view in mind, we may encode parsing algorithms and execute them entirely
internal to our formalism. Our contributions are then:

@@

Unfortunately, programming in a linear term language is often
cumbersome and unintuitive for users who are unaccustomed to the usage
restrictions.\max{You are arguing against our own type theory here...} \steven{I
  want to argue against a \emph{purely linear} calculus to motivate the
  convenience of LNL, but perhaps this isn't appropriate here, and I don't want to
  make our own type theory sound bad.Should probably just cut the beginning of
  this paragraph and jump into LNL, but then there's not an immediate reason for
why one might want/need it over a pure linear system} Code in such a language can
become unnecessarily verbose when ensuring every function written is indeed linear. It would be more ergonomic to allow users using Girard encodings to translate
high-level reasoning into low-level linear terms.\pedro{What encodings are these?
We should add a citation here}
To alleviate this
concern, in 1995 Benton et al.\ proposed an alternative categorical
semantics of linear logic based on adjoint interactions between linear
and non-linear logics \cite{bentonMixedLinearNonlinear1995} ---
appropriately referred to as a \emph{linear-non-linear} (LNL)
system. This work is simply typed, so the boundary between linear and
non-linear subtheories is entirely characterized via a monoidal
adjunction $F \dashv G$ between linear terms and non-linear terms.

@@

\pedro{The paragraphs above are a bit circular, what about the following alternative?}
\steven{I am in favor of cutting what's above, but we haven't introduced the
  syntax in the main body of the paper yet, even if the figures appeared
  earlier. So what you write needs to take that into account}

@@

\paragraph{Regular Expressions} We may realize regular expression as a subtheory
of our language by restricting the type constructors to the
linear unit, characters, $\otimes$, $\oplus$, and Kleene star. Classically, the
languages recognized by these are exactly the regular languages --- the lowest
on the Chomsky hierarchy.

@@

Though Kleene categories are expressive enough to reason about
concepts that are outside of reach of Kleene algebras, their
simply-typed nature makes them not so expressive from a type theoretic
point of view. This is limiting because type theories are successful
syntactic frameworks for manipulating complicated categorical
structures while avoiding some issues common in category theory, such
as coherence issues.

@@

\steven{Max suggests augmenting the definition of a Chomsky category to
  something like two categories $L$ and $U$, and $U$ is Cartesian closed I don't
  fully recall the rest.

@@

\pedro{This is likely true, but if we explicitly say so, this warrants a proof. I think that
if we don't say anything about the syntactic category, reviewers won't mind.}

\steven{I agree that we don't want to say anything that opens unnecessary
  questions for proofs we haven't written. However, it seems hard to make the
  case that we have the right categorical model of the syntax if this statement
  isn't true. By restating our definition of Chomsky category, this should be
  obvious or a quick proof}

@@

Further, semantic actions can be arranged into a structured category.
Define $\SemAct$ as the comma
category $\Grammar / \Delta$, where $\Delta : \Set \to \Grammar$ defines a
discrete presheaf. That is, for a set $X$, $\Delta (X)(w) = X$ for all
$w \in \String$. As $\SemAct$ is defined as a comma category, it has a forgetful
functor into $\Grammar$. That is, $\SemAct$ serves as a notion of formal
grammar. Moreover, $\SemAct$ is a model of \theoryabbv.

@@

The canonicity theorem from the previous section gives a tight connection
between terms in our type theory and the theory of formal languages. In this section
we further explore these connections by using the fact that, classically
formal language theory is closely related to the study of automata.
In the Chomsky hierarchy, each language class is associated to a class of
automata that serve as recognizers. Internal to \theoryabbv,
we can characterize these language classes syntactically; moreover, we
demonstrate the equivalence of these language classes to its associated automata class as a
proof term within our logic.

@@

In 1992, while investigating the addition of arbitrary recursion to Kleene
algebra \cite{leiss},
Lei{\ss} used a least fixed-point operator instead of the Kleene star. Although
he did not explicitly make use of a type system, or have access to the view of Frisch and
Cardelli, Lei{\ss} had added the full power of inductive types to a type system of regular expressions.

@@

\steven{Mention Neel's typed algebraic approach paper here. Say it is similar at
  first glance, as its parsing with types, but that he actually is representing
  something quite different. Or is it so different that its not worth mentioning?}

@@

\steven{Can we sketch here what an apt categorification of KATs might be? As a
  poor first pass: perhaps a Kleene category with a full involutive
  subcategory that corresponds to the Boolean algebra?.}
\steven{What can we say about internalizing Kleene algebra proofs or extensions?}
\steven{any KA citations needed above?}

@@

\cite{fu2023twolevellineardependenttype} give a compiler for a two-level
linear dependent type theory. This could serve as a template for a
compiler implementation of \theoryname that is not just inference rules
embedded in Agda.

\paragraph{Implementations in Parsing}
Talk about how we have a DFA parser in Agda and want to to get out a regex
parser. Then context free, etc

@@

\paragraph{Other Future Work}
Backtracking parsing: via (delimited) continuations? effect handlers?
